Honors Thesis Proposal

### 2009-03-13 honors_thesis_proposal/paper.tex
Renamed honors_thesis to honors_thesis_proposal

Optimization problems are ubiquitous in our modern world.  Businesses such as
Google need to decide where to place ads, scientists need to fit models to
data, and airlines need to schedule their flights.  All of those problems are
optimization problems, and optimization techniques can be used to find
solutions.  Optimization algorithms are derived from all sorts of natural
phenomena.  Some methods look at the landscape of the function as they are
searching and follow hills to the best values, as in the gradient ascent
method.  Other methods take ideas from biology or metallurgy, like genetic
algorithms and simulated annealing.  Particle swarm optimization is a recently
developed optimization technique that draws on ideas from the sociology of
flocking birds.  All of these methods, however, are fundamentally sequential in
nature---they need to be run in a specific order, and because of that, they are
typically only run on one machine.  The problems people want to solve are
getting bigger, and larger problems need to run on multiple machines in
parallel.  Parallel implementations of these algorithms have been attempted,
but they do not often perform as well as sequential versions for the amount of
computation performed.  The work I propose focuses on one of those algorithms,
particle swarm optimization, and improving its performance in a large-scale,
parallel world.

@@

The bulk of the work for this project will be writing code and running
experiments.  I have already read many conference papers on particle swarm
optimization to know the current state of the field, so I will know what the
results need to be compared to.  In implementing speculative execution, I will
take care that the algorithm behaves exactly like the original, standard
particle swarm optimization (PSO) algorithm in every aspect except the
speed-up.  The same constants will be used as are commonly published in the
literature.

In running experiments, many trials will be run and averaged, as there is a
random component in this algorithm.  I will use statistical tests for all
comparisons to the original PSO algorithm.  The success criterion will be a
better performance of the algorithm as measured by the lowest function
evaluation seen over a number of iterations for several benchmark and
real-world problems.  Speculative execution takes more computation than regular
PSO with the same number of particles, so the two algorithms will be compared
at points where they use equal amounts of processing power.

@@

The finished thesis will be presented as a conference paper, including the
following sections:

@@

Speculative execution in a program is the execution of code that may or may not
end up actually being needed.  Modern processors routinely do this when a
conditional branch is encountered; they try to predict which branch will
actually be needed, and continue their execution.  If it turns out that the
prediction was incorrect, the work is discarded.  But if the prediction was
right, the program can execute much faster than if it had waited on the branch.

In particle swarm optimization, speculative execution can be performed because
the next position a particle moves to doesn't depend on the value of the
function at its current position.  That means you can just compute all possible
next positions (depending on whether or not the gbest or pbest were updated---\gbest 
and \pbest in equation \ref{eq:velupdate} got updated), and evaluate each position 
in parallel.  That computes two iterations at the same time.

@@

I am a Computer Science major.  My main focuses in my undergraduate career have
been optimization, natural language processing, and machine learning.  The
optimization emphasis has come mainly in my work with Dr. Kevin Seppi, for whom
I am a research assistant.  I have spent the last year working with him,
working mainly on particle swarm optimization topics.  We worked together with
Dr. Branton Campbell, a physics professor, to solve a crystallography problem
for him using a parallel implementation of PSO.  Those results will be
submitted to Nature Materials.  Recently we, along with a graduate student who
works in the lab, submitted a paper to the Congress on Evolutionary
Computation, a major venue for optimization techniques.  The paper was on
topologies and communication in large particle swarms, and formed the
beginnings of the work I propose.

>>>>

Optimization problems are ubiquitous in our modern world.  Businesses such as
Google need to decide where to place ads, scientists need to fit models to
data, and airlines need to schedule their flights.  All of those problems are
optimization problems, and optimization techniques can be used to find
solutions.  Optimization algorithms have been derived from all sorts of natural
phenomena.  Some methods look at the landscape of the function as they are
searching and follow hills to the best values, as in the gradient ascent
method.  Other methods take ideas from biology or metallurgy, like genetic
algorithms and simulated annealing.  Particle swarm optimization is a recently
developed optimization technique that draws on ideas from the sociology of
flocking birds.  All of these methods, however, are fundamentally sequential in
nature---they need to be run in a specific order, and because of that, they are
typically only run on one machine.  The problems people want to solve are
getting bigger, and larger problems need to run on multiple machines in
parallel.  Parallel implementations of these algorithms have been attempted,
but they do not often perform as well as sequential versions for the amount of
computation performed.  The work I propose focuses on one of those algorithms,
particle swarm optimization, and improving its performance in a large-scale,
parallel world.

@@

The bulk of the work for this project will be writing code and running
experiments.  I have already read many conference papers on particle swarm
optimization to know the current state of the field, so I will know what the
results need to be compared to.  In implementing speculative execution, I will
take care that the algorithm behaves exactly like the original, standard
particle swarm optimization (PSO) algorithm in every aspect except the
speed-up.  My work can then be accurately compared to other work in the field.

In running experiments, many trials will be run and averaged, as there is a
random component in this algorithm.  I will use statistical tests for all
comparisons to the original PSO algorithm.  The success criterion will be a
better performance of the algorithm as measured by the lowest function
evaluation seen over a number of function evaluations for several benchmark and
real-world problems.  Speculative execution takes more computation than parallel
PSO with the same number of particles, so the two algorithms will be compared
at points where they use equal amounts of function evaluations.

@@

The finished thesis will be presented in the format of a peer-reviewed
conference paper, including the following sections:

@@

Speculative execution in a program is the execution of code that may or may not
end up actually being needed.  Modern processors routinely do this when a
conditional branch is encountered; they try to predict which branch will
actually be needed and continue their execution.  If it turns out that the
prediction was incorrect, the work is discarded.  But if the prediction was
right, the program can execute much faster than if it had waited on the branch.

In a lot of optimization algorithms, speculative execution is impossible,
because the next iteration of the algorithm depends on the evaluation of the
function during the previous iteration.  In simulated annealing, for instance,
the probability of accepting a new position depends on the difference in
function value between the new position and the current one.  In gradient
descent, the gradient of the function must be determined (which includes
actually evaluating it) to know how far to move before sampling the function
again.  But particle swarm optimization is rather unique in that it doesn't
require the evaluation of the function to know what the next position of each
particle will be.  It simply depends on \emph{which} particle ended up having
the best position and whether or not the current particle found a better spot
than it had seen before.  That means you can just compute all possible next
positions---all combinations of which particle is the new gbest and whether or
not pbest was updated (\gbest and \pbest from \eqref{eq:velupdate})---and
evaluate each position in parallel.  That computes two iterations at the same
time.  It uses a lot of extra work to get the second iteration, but for some
functions more iterations are worth more than extra particles.

@@

I am a Computer Science major.  My main focuses in my undergraduate career have
been optimization, natural language processing, and machine learning.  The
optimization emphasis has come mainly in my work with Dr. Kevin Seppi, for whom
I am a research assistant.  I have spent the last year working with him,
working mainly on particle swarm optimization topics.  We worked together with
Dr. Branton Campbell, a physics professor, to solve a crystallography problem
for him using a parallel implementation of PSO.  Those results will be
submitted to Nature Materials.  Recently we, along with Andrew McNabb, a
graduate student who works in the lab, submitted a paper to the Congress on
Evolutionary Computation, a major venue for optimization techniques.  The paper
was on topologies and communication in large particle swarms, and formed the
beginnings of the work I propose.
